---
title: "Performing differential circadian rhythm analysis using `mixedcirc`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mixedcirc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



## Rhythmicity analysis

In this part we do rhythmicity analysis on individual variables using the following model:

$$
expression = \sum_{i = 1}^{2}{group_{i}} + \\
\sum_{i=1}^{2}{group_{i} \times\cos\Big{(}\frac{2\times\pi\times time}{24}\Big{)}} + \\
\sum_{i=1}^{2}{group_{i} \times\sin\Big{(}\frac{2\times\pi\times time}{24}\Big{)}} + \\
random(subject)
$$

The evaluation of the model for global rhythmicity is performed using different statistics on combined null hypothesis:

$$
group_{i} \times cos\Big{(}\frac{2\times \pi \times time}{24}\Big{)} = 0 \\
group_{i} \times sin\Big{(}\frac{2\times \pi \times time}{24}\Big{)} = 0 \\
$$


The evaluation of differential rhythmicity is performed using different statistics on combined null hypothesis:

$$
group_{1} \times cos\Big{(}\frac{2\times \pi \times time}{24}\Big{)} - group_{2} \times cos\Big{(}\frac{2\times \pi \times time}{24}\Big{)} = 0 \\
group_{1} \times sin\Big{(}\frac{2\times \pi \times time}{24}\Big{)} - group_{2} \times sin\Big{(}\frac{2\times \pi \times time}{24}\Big{)} = 0 \\
$$

The amplitude is transformed using:

$$
amplitude_{\text{group}_{i}}=\sqrt{ \Big{(}group_{i} \times cos(\frac{2\times \pi \times time}{24})\Big{)}^2 + \Big{(}group_{i} \times sin(\frac{2\times \pi \times time}{24})\Big{)}^2 }
$$

The phase is calculated as:

          
$$
phase_{\text{group}_{i}} = \arctan{\frac{group_{i} \times sin(\frac{2\times \pi \times time}{24})}{group_{i} \times cos(\frac{2\times \pi \times time}{24})}}
$$
and further corrected:

$$
\DeclareMathOperator{\sign}{sign}
phase_{\text{group}_{i}} = \left\{
  \begin{array}{@{}ll@{}}
     - phase_{\text{group}_{i}}, & \text{if}\ (sb=1 \ \lor \ sb=0) \ \land \ sg=1 \\
    phase_{\text{group}_{i}}-\pi, & \text{if}\ sb=-1 \ \land \ (sg=0 \ \lor \ sg=1) \\
    -phase_{\text{group}_{i}}-\pi, & \text{if}\ (sb=-1 \ \lor \ sb=0) \ \land \ sg=-1\\
    phase_{\text{group}_{i}}-2\times \pi, & \text{if}\ sb=1 \ \land \ (sg=1 \ \lor \ sg=-1) 
  \end{array}\right.
$$
where $sb$ is defined as sign of $sb=cos(\frac{2\times \pi \times time}{24})$ and $sg$ as the sign of $sg=sin(\frac{2\times \pi \times time}{24})$. The sign is defined as:


$$
x \longmapsto \sign{x}=\left\{
  \begin{array}{@{}ll@{}}
    1, & \text{if}\ x\gt0 \\
    0, & \text{if}\ x= 0 \\
    -1, & \text{if}\ x\lt0 \\
  \end{array}\right.\\
$$
To avoid the negative phase, we then take the absolute value of the phase so $phase_{\text{group}_{i}}=|phase_{\text{group}{_i}}|$

## How to run the package

The main function of the package is `mixedcirc_detect`. It accepts a data frame of $N\times P$ and performs required analysis. 

For example, let's our simulated data be:

```{r setup,warning=FALSE,error=FALSE,message=FALSE}
library('mixedcirc')
data('circa_data')
str(circa_data)
```

This dataset contains a data matrix of $N = $`r nrow(circa_data$data_matrix)` samples and $P = $ `r nrow(circa_data$data_matrix)` variables. It also includes time, group and id for each of the samples. 

We can now do differntial rhythmicity analysis using:

```{r warning = FALSE, error = FALSE, message = FALSE}
results <- mixedcirc_detect(data_input = circa_data$data_matrix,
                            time = circa_data$time,
                            group = circa_data$group,
                            id = circa_data$id,
                            period = 24,
                            verbose = FALSE)
```

We can see that, in total, we have `r length(results)` variables. The result of each, can be printed as follows:

```{r,warning=FALSE,error=FALSE,message=FALSE}
print(results[1])
```
We can also plot an individual variable like this:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
plot(results[1])
```

The plotting function gives you several options to tune how the data is plotted. For example, one can add smoothing line to the data and check where the "data-driven" lines are following the fitted lines.

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
plot(results[1], plot_smooth=TRUE, plot_points=FALSE)
```

In the above plot, the dashed line (smoothing line) perfectly follows the solid line (the fit of the model), indicating that our fit is supported by the data.

You can also do significant test on individual fitted models:


```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
boot_results<-mixedcirc_boottest(results[1],nsim = 10,seed = 10)
print(boot_results)
```



## Data integration

`mixedcirc` provides support for doing integrative rhythm analysis (e.g., multiOmics) using canonical correlations. In this type of analysis, there are multiple types of data sets (e.g., RNA-seq and metabolomics) and the we are interested to extract a pattern that is supported by the majority of the datasets.

This functionality is provided by the `mixedcirc_integrate` function. To use this, we first need to prepare our data input (`data_input`).

Data input is a list of `data.frames` or matrices, each containing one of the omics/data sets.


```{r, warning=FALSE, error=FALSE, message=FALSE}
library('mixedcirc')
data('circa_data')
data_1 <- circa_data$data_matrix[, 1:3]
colnames(data_1) <- paste("data_1_", 1:ncol(data_1), sep = "")
data_2<-circa_data$data_matrix[, 3:7]
colnames(data_2) <- paste("data_2_", 1:ncol(data_2), sep = "")
data_input <- list(a=data_1, b=data_2)
lapply(data_input, dim)
```

In the code above, we have created two data sets, `a` and `b` and placed them inside a list. Please, note that we have created these artificial data by splitting our internal data sets. In reality, each of this data sets should be independent.

We can now simply, use the `mixedcirc_integrate` function to do the data integration.

```{r, warning=FALSE, error=FALSE, message=FALSE}
results <- mixedcirc_integrate(data_input, 
                               time = circa_data$time, 
                               group = circa_data$group, 
                               id = circa_data$id)
```

The result returned by this function is an object of `mixedcirc_integration` class that includes partial scores (for each data type), average scores and loadings. We can extract this information using `mixedcirc_getscore` and perform rhythm analysis using `mixedcirc_detect`. 

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
data_matrix <- mixedcirc_getscore(results, type = 'average')
fitted_data <- mixedcirc_detect(data_matrix, 
                                time = circa_data$time, 
                                group = circa_data$group, 
                                id = circa_data$id)
plot(fitted_data[1], plot_points=FALSE, ylab='L1')
```

The plotted scores are supposed to have captured the overall trend of the data. The loadings can be extracted and used to select variables or do clustering.

The package also outputs the original model that was generated by `mixOmics` package. This model can be used in `mixOmics` package to plot and do other manipulations. For example, we can plot the loadings for the data set `b` on component 1:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=5}
model <- mixedcirc_getscore(results, type = "model")
mixOmics::plotLoadings(model, block="b", size.title = .7)
```

This plot shows that the contribution of the `data_2_4` variable is the largest from all the variables. More importantly, it will tell us that `data_2_1` vs. `data_2_3` have more similar pattern compared to `data_2_2` vs. `data_2_3`

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
results <- mixedcirc_detect(
  data_input = data_input$b[, c("data_2_1","data_2_2","data_2_4"), drop=F], 
  time = circa_data$time,
  group = circa_data$group, id = circa_data$id,period = 24,
  verbose = FALSE)
cowplot::plot_grid(plot(results[1]), plot(results[2]), plot(results[3]))
```

## Detrending and heatmap

`mixedcirc` can also be used to remove linear trend in the data, making it easier to visualize the pattern using, for example, heatmaps.

Function `mixedcirc_detrend` fits a linear trend to the fitted cosine model and returns a matrix containing detrended data.

Let's, for the start, plot the heatmap of the data:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
library(mixedcirc)
data("circa_data")

my_data <- t(circa_data$data_matrix[order(circa_data$group, circa_data$time), ])
my_colors <- c("red","blue")[as.factor(circa_data$group[order(circa_data$group)])]
heatmap(x = my_data, 
        Rowv = NA, 
        Colv = NA,
        ColSideColors = my_colors)

```

The heatmap already looks very good (simulated data of course!). Let's do the same thing after de-trending.

We first fit our model using our usual functions:

```{r, warning=FALSE, error=FALSE, message=FALSE}
results <- mixedcirc_detect(data_input = circa_data$data_matrix,
                          time = circa_data$time,
                          group = circa_data$group,
                          id = circa_data$id,
                          period = 24,
                          verbose = FALSE)

```

We can now do the detrending directly on the `results` object:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
detrended <- mixedcirc_detrend(results,per_group = FALSE, verbose = FALSE)
my_data <- t(detrended[order(circa_data$group, circa_data$time), ])
my_colors <- c("red","blue")[as.factor(circa_data$group[order(circa_data$group)])]
heatmap(x = my_data, 
        Rowv = NA,
        Colv = NA,
        ColSideColors = my_colors)
```

We can see that the trend is slightly more visible. In fact, by performing detrending, we can remove all the visual artifacts that can, otherwise, negatively impact the heatmap or similar visualizations.
