---
title: "Performing circadian and differential circadian rhythm analysis using `mixedcirc`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Performing circadian and differential circadian rhythm analysis using mixedcirc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `mixedcirc` package provides functions for circadian rhythm analysis using linear models and mixed-effects models. The main function, `mixedcirc_detect()`, fits one model per variable and returns estimates of mesor, amplitude, phase, and p-values for rhythmicity. When group information is available, the function can also test for differences in rhythmicity between groups.

The current version of `mixedcirc_detect()` expects:

* `data_input`: a matrix, data.frame, or `DGEList` containing the measured variables
* `meta_input`: a data.frame containing the metadata
* `time`, `group`, `id`, and `replicate_id`: character strings pointing to column names in `meta_input`

## Circadian model

For each variable, circadian time is represented using cosine and sine terms:

$$
\text{inphase} = \cos\left(\frac{2\pi \times time}{period}\right)
$$

$$
\text{outphase} = \sin\left(\frac{2\pi \times time}{period}\right)
$$

For grouped non-RRBS data, the default fixed-effects model is:

$$
\text{measure} \sim 1 + group + group:\text{inphase} + group:\text{outphase}
$$

If a mixed-effects model is used, a random-effect term is added depending on the selected `random_effect_type`. For example, with a random intercept by subject, the model becomes conceptually:

$$
\text{measure} \sim 1 + group + group:\text{inphase} + group:\text{outphase} + (1 \mid subject)
$$

If no group variable is provided, all samples are treated as belonging to a single group and the default model becomes:

$$
\text{measure} \sim 1 + \text{inphase} + \text{outphase}
$$

## Rhythmicity and differential rhythmicity tests

For each group, rhythmicity is assessed by testing the joint null hypothesis:

$$
\beta_{\text{inphase}} = 0
$$

$$
\beta_{\text{outphase}} = 0
$$

That is, a variable is considered rhythmic only if at least one of the sine or cosine coefficients is different from zero.

When multiple groups are present, `mixedcirc_detect()` also performs pairwise tests for differential rhythmicity. Conceptually, for two groups, this corresponds to testing:

$$
\beta_{\text{inphase},1} - \beta_{\text{inphase},2} = 0
$$

$$
\beta_{\text{outphase},1} - \beta_{\text{outphase},2} = 0
$$

In addition, a global rhythmicity test across all groups is reported.

## Amplitude and phase

For each group, amplitude is calculated from the cosine and sine coefficients as:

$$
\text{amplitude} = \sqrt{\beta_{\text{inphase}}^2 + \beta_{\text{outphase}}^2}
$$

Phase is derived from the estimated cosine and sine coefficients using the internal `calc_phase()` function. In practice, `mixedcirc` uses a quadrant-corrected arctangent-like calculation based on the signs of the estimated `inphase` and `outphase` coefficients.

If `abs_phase = TRUE`, which is the default, the absolute value of the estimated phase is returned.

## Example data

We first load the package and inspect the example data:

```{r setup, warning=FALSE, error=FALSE, message=FALSE}
library(mixedcirc)
```

We are now going to simulate some data using `mixedcirc_simulate_data` function. `mixedcirc_simulate_data` can be used to generate continuous or RRBS-like circadian data with optional covariates and optional lagged relationships between selected features.

For this example we are going to generate circedian data for two groups:

```{r, warning=FALSE, error=FALSE, message=FALSE}
sim_data <- mixedcirc_simulate_data(
  n_subjects_per_group = 10,
  n_timepoints = 8,
  n_features = 5,
  groups = c("A", "B"),
  seed = 1
)
```

`mixedcirc_simulate_data` returns a list containing four data frames.
```{r, warning=FALSE, error=FALSE, message=FALSE}
str(sim_data)
```

In order to use `mixedcirc_detect()` to perform rhythmicity analysis we need two data.frames: `data_input` which is simulated feature matrix and `meta_input` which  contains meta data.


`data_input` must be a numeric matrix, data.frame, or 'DGEList' containing the input data. Rows correspond to observations and columns correspond to variables.  

```{r, warning=FALSE, error=FALSE, message=FALSE}

head(sim_data$data_input)
```


`meta_input` is a data.frame or matrix containing the sample-level metadata used to build the design matrix. This must include the columns referenced by 'time', 'group', 'id', 'replicate_id', and any terms in 'formula_extra'.

```{r, warning=FALSE, error=FALSE, message=FALSE}

head(sim_data$meta_input)
```

## Running circadian analysis

We can now perform circadian and differential circadian analysis using `mixedcirc_detect()`:

```{r, warning=FALSE, error=FALSE, message=FALSE}
results <- mixedcirc_detect(
  data_input = sim_data$data_input,
  meta_input = sim_data$meta_input,
  time = "time",
  group = "group",
  id = "id",
  period = 24,
  verbose = FALSE
)
```

The result is a `mixedcirc_fit_list` object containing one fitted result per variable:

```{r, warning=FALSE, error=FALSE, message=FALSE}
length(results)
```

We can inspect the first fitted result:

```{r, warning=FALSE, error=FALSE, message=FALSE}
print(results[1])
```

## Plotting individual fitted variables

An individual fitted variable can be plotted directly:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
plot(results[1])
```

The plotting function provides additional options. For example, we may overlay a smooth curve and hide the points:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
plot(results[1], plot_smooth = TRUE, plot_points = FALSE)
```

The solid line shows the fitted circadian model, while the dashed smooth curve provides a data-driven summary of the observed pattern.

## Bootstrap testing for an individual fitted model

You can also perform a bootstrap-based significance test on an individual fitted model:

```{r, warning=FALSE, error=FALSE, message=FALSE}
boot_results <- mixedcirc_boottest(results[1], nsim = 10, seed = 10)
print(boot_results)
```

## Using additional covariates

The function allows you to include additional fixed effects through `formula_extra`. These covariates must be present in `meta_input`.

For example, let us create an artificial batch and age variables and include them in the model:

```{r, warning=FALSE, error=FALSE, message=FALSE}
sim_data_cov <- mixedcirc_simulate_data(
  n_subjects_per_group = 10,
  n_timepoints = 8,
  n_features = 5,
  groups = c("A", "B"),
  seed = 1,
  include_covariates = TRUE,
  batch_effects=c(B1=0,B2=10)
)

```

We now fit two models, one with and one without covariate adjustment and compare the results:

```{r, warning=FALSE, error=FALSE, message=FALSE}

results_without_adjustment <- mixedcirc_detect(
  data_input = sim_data_cov$data_input,
  meta_input = sim_data_cov$meta_input,
  time = "time",
  group = "group",
  id = "id",
  period = 24,
  verbose = FALSE
)

results_with_adjustment <- mixedcirc_detect(
  data_input = sim_data_cov$data_input,
  meta_input = sim_data_cov$meta_input,
  time = "time",
  group = "group",
  id = "id",
  period = 24,
  verbose = FALSE,formula_extra = ~batch+age
)


plot(results_without_adjustment[1],plot_points=F,plot_title="without adjustment")
plot(results_with_adjustment[1],plot_points=F,plot_title="with adjustment")

```

You can see the effect of addition covariates have been removed, making the fitted curve focusing on the rhythm. 

## Comparing two rhythms

It often happens that we have multiple rhythms and would like to know whether there is an association between their curves. This can be done by comparing the curves directly or by performing a lag analysis. Lag analysis examines whether one rhythm is shifted in time relative to another. In other words, it asks whether changes in one curve tend to occur before, after, or at the same time as changes in the other curve. This can help identify possible delays, phase shifts, or lead–lag relationships between the rhythms.

`mixedcirc` supports cross correlation and association using `mixedcirc_compare` function. To demonstrate it let's simulate data with a known lag:

```{r, warning=FALSE, error=FALSE, message=FALSE}
sim_data_lag <- mixedcirc_simulate_data(
  n_subjects_per_group = 10,
  n_timepoints = 8,
  n_features = 5,
  groups = c("A", "B"),
  cross_lag_spec = list(
    list(from = 1, to = 2, lag = 1, beta = 1.0),
    list(from = 3, to = 4, lag = c(A = 1, B = 2), beta = 0.7)
  ),
  seed = 1
)


```
Here using `cross_lag_spec` parameter we introduced two lagged correlation. Feature 2 is a lagged version of feature 1 and feature 4 is the lagged version of feature 3. 

We can now fit the data using `mixedcirc_detect`:


```{r, warning=FALSE, error=FALSE, message=FALSE}
results_lag <- mixedcirc_detect(
  data_input = sim_data_lag$data_input,
  meta_input = sim_data_lag$meta_input,
  time = "time",
  group = "group",
  id = "id",
  period = 24,
  verbose = FALSE
)

plot(results_lag[3],plot_points=F,plot_title="feature 3")
plot(results_lag[4],plot_points=F,plot_title="feature 4")
```

and now we can compare the fitted models. For example we can compare feature 3 and 4 using `mixedcirc_compare` for lags from -2 to 2:

```{r, warning=FALSE, error=FALSE, message=FALSE}
mixedcirc_compare(results_lag[3],results_lag[4],lags = -2:2)
```


You can see that for this comparison the method correctly identified the lag at 1 and 2 unit shifts. 


## Detrending and heatmaps

`mixedcirc` can also be used to remove linear trends from fitted variables, which may improve downstream visualization such as heatmaps.

The `mixedcirc_detrend()` function operates on fitted `mixedcirc` results and returns a detrended matrix.

We first visualize the raw data:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}

sim_data <- mixedcirc_simulate_data(
  n_subjects_per_group = 10,
  n_timepoints = 8,
  n_features = 50,
  groups = c("A", "B"),
  seed = 1
)

results <- mixedcirc_detect(
  data_input = sim_data$data_input,
  meta_input = sim_data$meta_input,
  time = "time",
  group = "group",
  id = "id",
  period = 24,
  verbose = FALSE
)

res_tab <- do.call(rbind, lapply(seq_along(results), function(i) as.data.frame(results[i]@results)))
ord_features <- order(res_tab$phase_A, res_tab$amps_A, na.last = TRUE)
sample_ord <- order(sim_data$meta_input$group, sim_data$meta_input$time)

my_data <- t(sim_data$data_input[sample_ord, ])
my_data <- my_data[ord_features, ]
my_colors <- c("red", "blue")[as.factor(sim_data$meta_input$group[sample_ord])]

heatmap(
  x = my_data,
  Rowv = NA,
  Colv = NA,
  ColSideColors = my_colors
)
```

Now we fit the model:

```{r, warning=FALSE, error=FALSE, message=FALSE}

```

We can then detrend the fitted data:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
detrended <- mixedcirc_detrend(results, per_group = T, verbose = FALSE)

my_data <- t(detrended[sample_ord, ])
my_data <- my_data[ord_features, ]
my_colors <- c("red", "blue")[as.factor(sim_data$meta_input$group[sample_ord])]

heatmap(
  x = my_data,
  Rowv = NA,
  Colv = NA,
  ColSideColors = my_colors
)

```

After detrending, rhythmic patterns may become easier to see because non-circadian trends are reduced.


## Data integration

`mixedcirc` also provides experimental support for integrative rhythm analysis across multiple datasets, for example when combining transcriptomics and metabolomics. In that setting, the goal is to extract components or scores that capture patterns shared across several data types.

This functionality is provided by `mixedcirc_integrate()`. The input should be a list of matrices or data.frames, one for each dataset.

Below we create two small example datasets by splitting the internal example matrix:

```{r, warning=FALSE, error=FALSE, message=FALSE}

sim_data <- mixedcirc_simulate_data(
  n_subjects_per_group = 10,
  n_timepoints = 8,
  n_features = 50,
  groups = c("A", "B"),
  seed = 1
)




data_1 <- sim_data$data_input[, 1:25, drop = FALSE]
colnames(data_1) <- paste0("data_1_", seq_len(ncol(data_1)))

data_2 <- sim_data$data_input[, 26:50, drop = FALSE]
colnames(data_2) <- paste0("data_2_", seq_len(ncol(data_2)))

data_input <- list(a = data_1, b = data_2)
lapply(data_input, dim)
```

We can then integrate the datasets:

```{r, warning=FALSE, error=FALSE, message=FALSE}
integration_results <- mixedcirc_integrate(
  data_input,meta_input = sim_data$meta_input,
  time = "time",
  group ="group",
  id = "id"
)
```

The returned object contains integrated scores and loadings. These can be extracted using `mixedcirc_getscore()`.

For example, we can extract the average scores and analyze them with `mixedcirc_detect()`:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=7}
score_matrix <- mixedcirc_getscore(integration_results, type = "average")

fitted_scores <- mixedcirc_detect(
  data_input = score_matrix,
  meta_input = sim_data$meta_input,
  time = "time",
  group = "group",
  id = "id",
  period = 24,
  verbose = FALSE
)

plot(fitted_scores[1], plot_points = FALSE, ylab = "L1")
```

The integrated scores are expected to summarize the shared temporal trend across the datasets.

The original integration model can also be extracted and reused for additional visualizations. For example, we can visualize the loadings for block `b`:

```{r, warning=FALSE, error=FALSE, message=FALSE, fig.height=5, fig.width=5}
model <- mixedcirc_getscore(integration_results, type = "model")
mixOmics::plotLoadings(model, block = "b", size.title = 0.7)
```

This helps identify variables that contribute strongly to the integrated component.


## Notes

A few important practical notes:

* `time`, `group`, `id`, and `replicate_id` are column names in `meta_input`, not raw vectors.
* If `id` is not provided, `mixedcirc_detect()` switches to ordinary linear regression.
* If `RRBS = TRUE`, the data are assumed to be arranged as methylated and unmethylated rows for each replicate, and `replicate_id` must be provided.
* Additional covariates can be added through `formula_extra`.
* No multiple-testing correction is performed automatically inside `mixedcirc_detect()`.

## Summary

The `mixedcirc` package provides a flexible framework for:

* circadian rhythm analysis of individual variables
* differential rhythmicity analysis between groups
* adjustment for additional covariates
* integration of multiple omics datasets
* detrending and visualization of rhythmic signals

The main workflow is centered around `mixedcirc_detect()`, which returns fitted models together with interpretable rhythm summaries such as mesor, amplitude, phase, and p-values.


